---
title: "Reproducible Research: Peer Assessment 1"
output: 
        html_document:
                keep_md: true
---

Updates as of `r format(Sys.time(), "%H:%M, %a, %b %d (%Y)")`:

(1) Fixed a few typos and minor grammar mistakes.

(2) Rename the data.frame object `r "ave3"` as `r "aves"` to distinguish from 
the two numeric vectors `r "ave"` & `r "ave2"`.

## Loading and preprocessing the data
- Load the data. You can download `r "repdata-data-activity.zip"` from the link 
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).
```{r, echo = TRUE, cache = TRUE}
link <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zip <- "repdata-data-activity.zip"
if (!file.exists("activity.csv")) {
        if (!file.exists(zip)) {
                download.file(url = link, destfile = zip)
                unzip(zip)
        } else {
                unzip(zip)
        }
}
```
- Load the data and process the data into a data frame.
```{r}
cols <- c("numeric", "POSIXct", "numeric")
dat <- read.csv("activity.csv", na.strings = c("NA", ""), colClasses = cols)
```

## What is the mean total number of steps taken per day?
- Calculate the total number of steps taken each day.
```{r}
sums <- aggregate(steps ~ date, data = dat, sum, na.rm = TRUE)
```
- Make a histogram of the total steps taken each day.
```{r histogram1}
library(datasets)
hist(sums$steps, col = "burlywood2", border = "burlywood4", breaks = 20,
     main = "Histogram of the Total Steps Taken Each Day", 
     xlab = "Number of Steps")
```

- Calculate and report the mean of "the total steps taken per day".
```{r}
ave <- mean(sums$steps, na.rm = TRUE)
print(ave)
```
The mean of the total steps is `r format(ave, scientific = FALSE)` steps per 
day.

- Calculate and report the median of "the total steps taken per day".
```{r}
med <- median(sums$steps, na.rm = TRUE)
print(med)
```
The median of the total steps is `r format(med, scientific = FALSE)` steps per 
day.

## What is the average daily activity pattern?
- Calculate the mean steps in each 5-minute interval averaged across all days.
```{r}
daily <- aggregate(steps ~ interval, data = dat, mean, na.rm = TRUE)
```
- Convert intervals to proper time format. (1970-01-01 is chosen for 
convenience in later calculations only)
```{r}
hr <- daily$interval %/% 100
min <- daily$interval %% 100
daily$interval <- as.POSIXct(paste("1970-01-01", hr, min), 
                             format = "%Y-%m-%d %H %M")
```
- Make a time series plot of the 5-minute interval (x-axis) and the mean steps 
taken, averaged across all days (y-axis).
```{r xyplot1}
locs <- c("1970-01-01 00:00:00", "1970-01-01 06:00:00", "1970-01-01 12:00:00", 
              "1970-01-01 18:00:00", "1970-01-02 00:00:00")
ticklocs <- as.POSIXct(locs, format = "%Y-%m-%d %H:%M:%S")
ticknames <- c("00:00", "06:00", "12:00", "18:00", "24:00")
library(lattice)
xyplot(steps ~ interval, data = daily, type = "l", xlab = "Interval", 
       ylab = "Number of Steps", 
       main = "Time Series Plot of the Daily Activity Pattern", 
       scales = list(x = list(at = ticklocs, labels = ticknames)))
```

- Find the 5-minute interval that, on average across all the days in the dataset, 
contains the maximum number of steps.
```{r}
MAXstp <- max(daily$steps, na.rm = TRUE)
MAXint <- daily[which(daily[,"steps"] == MAXstp), "interval"]
substr(MAXint, 12, 16)
```
The 5-minute interval starting at `r substr(MAXint, 12, 16)` on average across 
all the days, contains the maximum number of steps.

## Imputing missing values
- Calculate and report the total number of rows with missing values.
```{r}
NArows <- sum(!complete.cases(dat))
print(NArows)
```
The total number of rows with missing values is `r NArows`.

- Create an updated dataset with all NAs filled in. 
(Fill in each missing value with the mean for that 5-minute interval)
```{r, cache = TRUE}
dat2 <- dat
for (i in 1:nrow(dat2)) {
        if (is.na(dat2$steps[i])) {
                if (i%%288 == 0) {
                        dat2$steps[i] <- daily$steps[288]
                } else {
                        dat2$steps[i] <- daily$steps[i%%288]
                }
        }
}
```
- Calculate the new total number of steps taken each day.
```{r}
sums2 <- aggregate(steps ~ date, data = dat2, sum)
```
- Make a new histogram of the total steps taken each day.
```{r histogram2}
library(datasets)
hist(sums2$steps, col = "darkseagreen2", border = "darkseagreen4", breaks = 20,
     main = "New Histogram of the Total Steps Taken Each Day", 
     xlab = "Number of Steps")
```

- Calculate and report the new mean of "the total steps taken per day".
```{r}
ave2 <- mean(sums2$steps)
print(ave2)
```
The new mean of the total steps is `r format(ave2, scientific = FALSE)`.

- Calculate and report the new median of "the total steps taken per day".
```{r}
med2 <- median(sums2$steps)
print(med2)
```
The new median of the total steps is `r format(ave2, scientific = FALSE)` steps 
per day.

- Do these values differ from the estimates from the first part of this report?

The mean (`r format(ave2, scientific = FALSE)`) remains the same as the estimate 
from the first part of this report, 
but the median (`r format(med2, scientific = FALSE)`) differs slightly from the 
original estimate (`r format(med, scientific = FALSE)`).

- What is the impact of imputing missing data on the estimates of the total daily 
number of steps?

I think the purpose of trying to imput missing data is to make the dataset more 
complete. As shown in this report, the filling-in of missing values has changed 
the median of total steps taken per day, though the difference between the 
original median and the new median is small. In a much larger dataset, the 
presence of missing data could contribute to a conclusion that is more deviated 
from the reality, thus leading to mistakes in making decisions based on the 
dataset. Filling in these missing values with possible data would give us a 
more complete picture of the real situations.

## Are there differences in activity patterns between weekdays and weekends?
- Create a new factor variable in the dataset with two levels (weekday & 
weekend) indicating whether a given date is a weekday or weekend day.
```{r}
wkd <- weekdays(dat2$date)
dat2 <- cbind(dat2, wkd)
```
- Set the system locale to English (US) if it is not.
```{r}
if (!Sys.getlocale(category = "LC_TIME") == "English_United States.1252") {
        invisible(Sys.setlocale(category = "LC_TIME", "English"))
}
```
- Convert the interval in the new dataset to time format.
```{r}
dat2$interval <- daily$interval
```
- Create a new factor column in the new dataset indicating which day in a week 
the date on the same row is; then use it to sort the dataset based on whether 
the date is a weekday or a weekend day.
```{r}
wkd <- weekdays(dat2$date)
dat2 <- cbind(dat2, wkd)
dat3 <- split(dat2, dat2$wkd)
wkdy <- rbind(dat3[[1]], dat3[[2]], dat3[[5]], dat3[[6]], dat3[[7]])
wkdy$wkd <- "weekday"
wknd <- rbind(dat3[[3]], dat3[[4]])
wknd$wkd <- "weekend"
dat3 <- rbind(wkdy, wknd)
```
- Make a panel plot containing a time series plot of the 5-minute interval 
(x-axis) and the average number of steps taken, averaged across all weekday days 
or weekend days (y-axis).
```{r xyplot2}
aves <- aggregate(steps ~ interval + wkd, data = dat3, mean)
library(lattice)
xyplot(steps ~ interval | wkd, data = aves, type = "l", layout = c(1, 2), 
       xlab = "Interval", ylab = "Number of Steps", 
       main = "Time Series Plot of the Activity Patterns", 
       scales = list(x = list(at = ticklocs, labels = ticknames)))
```

Based on the panel plot generated above, it is easy to tell that there are 
some differences in activity patterns between weekdays and weekends mainly 
during the hours from 6 am to midnight.